Variational theory of non-relativistic quantum electrodynamics

### 2018-11-14 submit/main.tex
Edits during submission.

 In this Letter, we establish a variational framework to analyze complex light-matter systems from first principles. Although  \textit{ab initio} methods such as QEDFT are exact in principle and provide access to all observables, a number of practical difficulties arise related to: the lack of simple exchange-correlation functionals to describe the ground state energy, as well as other more involved observables, the difficulty of obtaining real-space information about the photons as they are affected by light-matter coupling, the difficulty of handling excited state energies, and the common use of the long-wavelength (dipole) approximation. A variational framework, as we show, flexibly allows a real-space description of the electrons and photons as they are modified by the coupling and also beyond the dipole approximation. Beyond these advantages, a variational framework also allows conceptual insights, as we shall show, into a simple non-perturbative theory of Lamb shifts, into a quasiparticle description of QED systems, and into the notion of Casimir forces in the limit of one atom. A variational framework also allows compact semi-analytical formulae to describe complex systems which may assist the development of functionals for use in QEDFT. 
 %In the nonperturbative regime of quantum electrodynamics, there are very few models which can be analytically solved. In particularly, the only exactly solvable models in all regimes of coupling parameters include the Rabi model~\cite{braak2011}, which describes the coupling of a two-level atom to a single mode, the Hopfield model \cite{hopfield1958theory}, which describes the coupling of a harmonic oscillator to any number of cavity modes, and Dicke models based on two-level systems and harmonic oscillators \cite{dicke1954coherence}. Any realistic system which deviates from these simpler models require numerical diagonalization approaches for an exact solution, which become unwieldy when the dimension of the electronic Hilbert space becomes large, as is the case when describing an electronic system in real space. It also becomes unwieldy when one needs to keep track of many cavity modes which can also be populated with many photons. Exciting new developments towards ab-initio studies of QED have proceeded by proposing reduced-quantity descriptions such as density-functional descriptions of the light-matter coupling \cite{ruggenthaler2014,pellegrini2015,flick2015,dimitrov2017,flick2018,flick2018b,schaefer2018}. With such density-functional descriptions, workers in this field have been able to compute ground-state and time-dependent properties of realistic molecules in optical cavities \cite{flick2017c}.

@@

Immediately, we notice that the term in the interaction Hamiltonian linear in the vector potential (the "$A\cdot p$ term") makes no contribution to the expectation value of the ground state of the energy in this ansatz. At second order in the $A\cdot p$ term, virtual photon processes arise, such as Lamb shifts, whose emitter-position-dependence gives rise to van der Waals and Casimir-Polder forces  \cite{scheel2009macroscopic}. Thus, we seek to capture the effect of this term. Physically, this term will mix the factorizable ground state of Eq.~(\ref{eq:ansatz}) with states that simultaneously have virtual excitations of the matter and the electromagnetic field. The resulting state is now non-factorizable and we thus conclude that the term in the Hamiltonian linear in the vector potential leads to \textit{correlations} in the system, and contributes wholly at lowest order to the correlation energy of QED ground and excited states. We note that correlations can also be included in the energy shifts of excited emitter states, as well as states that already have photonic excitations \footnote{We briefly note that this behavior of the $A\cdot p$ and $A^2$-term is similar to the ${r}\cdot D$ and $r^2$ term in the length-gauge reported in recent work on the optimized effective potential~\cite{pellegrini2015,flick2017c} method for QEDFT including one-photon processes. }.
\begin{figure}[t]
\includegraphics[width=8.5cm]{Figure2and3combined.pdf}
\caption{\textbf{Variational theory of ground and excited states in the ultra-strong coupling regime of QED.} (a) Lowest few energy levels of a two (top), three (middle), and four (bottom) level system embedded in the middle of a one-dimensional cavity. The results of our variational method (blue) are compared to perturbation theory (orange), as well as numerical diagonalization (red) with the Fock space truncated to fifty cavity modes with no more than four photons. (Inset) The fourth and fifth energy levels show a weak anti-crossing behavior which is well-reproduced by the variational theory. Blue denotes variational while red denotes numerical. (b) Mechanism of overestimation of couplings and resonances in perturbation theory: the modes derived from the variational theorem are always suppressed in the vicinity of the emitter, and this self-consistently decreases the coupling between the emitter and the field.  }
\label{fig:results}
\end{figure}

@@

\section{Acknowledgements}
We thank Prof. Joel Yuen-Zhou (University of California San Diego), Prof. Ido Kaminer (Technion Israel Institute of Technology), Prof. Marin Solja\v{c}i\'{c} (Massachusetts Institute of Technology) and Prof. John D. Joannopoulos (Massachusetts Institute of Technology), for useful discussions. N. R. recognizes the support of the DOE Computational Science Graduate Fellowship (CSGF) fellowship no.  DE-FG02-97ER25308. J. F. acknowledges financial support from the Deutsche Forschungsgemeinschaft (DFG Forschungsstipendium FL 997/1-1). This work was supported by the DOE Photonics at Thermodynamic Limits Energy Frontier Research Center under grant no. DE-SC0019140.

>>>>

In this Letter, we establish a variational framework to analyze complex light-matter systems from first principles. Although  \textit{ab initio} methods such as QEDFT are exact in principle and provide access to all observables, a number of practical difficulties arise related to: the lack of simple exchange-correlation functionals to describe the ground state energy, as well as other more involved observables, the difficulty of obtaining real-space information about the photons as they are affected by light-matter coupling, the difficulty of handling excited state energies, and the common use of the long-wavelength (dipole) approximation. A variational framework, as we shall show, flexibly allows a real-space description of the electrons and photons as they are modified by the coupling and also beyond the dipole approximation. Beyond these advantages, a variational framework also allows conceptual insights, as we shall show, into a simple non-perturbative theory of Lamb shifts, into a quasiparticle description of QED systems, and into the notion of Casimir forces in the limit of one atom. A variational framework also allows compact semi-analytical formulae to describe complex systems which may assist the development of functionals for use in QEDFT. 
 %In the nonperturbative regime of quantum electrodynamics, there are very few models which can be analytically solved. In particularly, the only exactly solvable models in all regimes of coupling parameters include the Rabi model~\cite{braak2011}, which describes the coupling of a two-level atom to a single mode, the Hopfield model \cite{hopfield1958theory}, which describes the coupling of a harmonic oscillator to any number of cavity modes, and Dicke models based on two-level systems and harmonic oscillators \cite{dicke1954coherence}. Any realistic system which deviates from these simpler models require numerical diagonalization approaches for an exact solution, which become unwieldy when the dimension of the electronic Hilbert space becomes large, as is the case when describing an electronic system in real space. It also becomes unwieldy when one needs to keep track of many cavity modes which can also be populated with many photons. Exciting new developments towards ab-initio studies of QED have proceeded by proposing reduced-quantity descriptions such as density-functional descriptions of the light-matter coupling \cite{ruggenthaler2014,pellegrini2015,flick2015,dimitrov2017,flick2018,flick2018b,schaefer2018}. With such density-functional descriptions, workers in this field have been able to compute ground-state and time-dependent properties of realistic molecules in optical cavities \cite{flick2017c}.

@@

Immediately, we notice that the term in the interaction Hamiltonian linear in the vector potential (the `$A\cdot p$ term') makes no contribution to the expectation value of the ground state of the energy in this ansatz. At second order in the $A\cdot p$ term, virtual photon processes arise, such as Lamb shifts, whose emitter-position-dependence gives rise to van der Waals and Casimir-Polder forces  \cite{scheel2009macroscopic}. Thus, we seek to capture the effect of this term. Physically, this term will mix the factorizable ground state of Eq.~(\ref{eq:ansatz}) with states that simultaneously have virtual excitations of the matter and the electromagnetic field. The resulting state is now non-factorizable and we thus conclude that the term in the Hamiltonian linear in the vector potential leads to \textit{correlations} in the system, and contributes wholly at lowest order to the correlation energy of QED ground and excited states. We note that correlations can also be included in the energy shifts of excited emitter states, as well as states that already have photonic excitations \footnote{We briefly note that this behavior of the $A\cdot p$ and $A^2$-term is similar to the ${r}\cdot D$ and $r^2$ term in the length-gauge reported in recent work on the optimized effective potential~\cite{pellegrini2015,flick2017c} method for QEDFT including one-photon processes. }.
\begin{figure}[t]
\includegraphics[width=8.5cm]{Figure2and3combined.pdf}
\caption{\textbf{Variational theory of ground and excited states in the ultra-strong coupling regime of QED.} (a) Lowest few energy levels of a two (top), three (middle), and four (bottom) level system embedded in the middle of a one-dimensional cavity. The results of our variational method (blue) are compared to perturbation theory (orange), as well as numerical diagonalization (red) with the Fock space truncated to fifty cavity modes with no more than four photons. (Inset) The fourth and fifth energy levels show a weak anti-crossing behavior which is well-reproduced by the variational theory. Blue denotes variational while red denotes numerical. (b) Mechanism of overestimation of couplings and resonances in perturbation theory: the modes derived from the variational theorem are always suppressed in the vicinity of the emitter, and this self-consistently decreases the coupling between the emitter and the field.  }
\label{fig:results}
\end{figure}

@@

%\section{Acknowledgements} We thank Prof. Joel Yuen-Zhou (University of California San Diego), Prof. Ido Kaminer (Technion Israel Institute of Technology), Prof. Marin Solja\v{c}i\'{c} (Massachusetts Institute of Technology) and Prof. John D. Joannopoulos (Massachusetts Institute of Technology), for useful discussions. N. R. recognizes the support of the DOE Computational Science Graduate Fellowship (CSGF) fellowship no.  DE-FG02-97ER25308. J. F. acknowledges financial support from the Deutsche Forschungsgemeinschaft (DFG Forschungsstipendium FL 997/1-1). This work was supported by the DOE Photonics at Thermodynamic Limits Energy Frontier Research Center under grant no. DE-SC0019140.

%We also acknowledge support from the STC Center for Integrated Quantum Materials NSF Grant No. DMR-1231319 and from the Harvard John A. Paulson School of Engineering and Applied Sciences.
